---
title: "Chapter 3 -  Data visualisation"
author: "Salman Mohammed"
date: "February 7, 2017"
output: html_document
---

#### Importing the necessary libraries
```{r results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
```


#### First look at the dataset
```{r}
mpg
```


#### A car's fuel efficiency vs. engine size
We want to see the relationship between a car's engine size (displ) and it's fuel efficiency
on the highway (hwy). A car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
```

In this plot, we can see a downward trend that explains a negative relationship between the two variables, so our hypothesis is that cars with bigger engines are less fuel efficient.

#### 3.2.4 - Exercises

1. Run ggplot(data = mpg) what do you see?
```{r}
ggplot(data = mpg)
```

2. How many rows are in mtcars? How many columns?
```{r}
nrow(mtcars)
ncol(mtcars)
```

3. What does the drv variable describe? Read the help for ?mpg to find out.
- drv describes the transmission system of the vehicle. The options are f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.
- hwy is the highway miles per gallon
- cyl is the number of cylinders
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = cyl, y =hwy))
```

5. What happens if you make a scatterplot of class vs drv. - class is the 'type' of car
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = class, y =drv))
```

6. Why is the plot not useful?
- The 'type' of car (e.g. minivan, 2-seater) does not determine the transmission system of the vehicle (e.g. front-wheel drive).




